---
title: "H~2~O Auto-Encoder Example"
output: 
     html_notebook
     highlight: pygments
     theme: spacelab
     toc: yes
---

https://shiring.github.io/machine_learning/2017/05/01/fraud
https://shiring.github.io/machine_learning/2017/05/02/fraud_2

```{r echo=FALSE}
library(tidyverse)
# download from https://www.kaggle.com/mlg-ulb/creditcardfraud - it is large!
creditcard <- read.csv(file.choose())
```
#Use Case

Autoencoders and anomaly detection: Use autoencoders to pre-train a classification modeland measure model performance on unbalanced data.

# The Data

The dataset gives > 280,000 instances of credit card use and for each transaction, we know whether it was fraudulent or not.  Datasets like this needs special treatment when performing machine learning because they are severely unbalanced: in this case, only 0.17% of all transactions are fraudulent.

While we could try to work with classifiers, like random forests or support vector machines, by applying over- or under-sampling techniques, we can alternatively try to find anomalies in the data (assuming we expect our fraud cases to be anomalies within the whole dataset).

When dealing with such a severe unbalance of response labels, we also need to be careful when measuring model performance. Because there are only a handful of fraudulent instances, a model that predicts everything as non-fraud will already achieve a > 99% accuracy. But despite its high accuracy, such a model won’t necessarily help us find fraudulent cases - the proverbial “needle-in-a-haystack” - that we actually want to find!

# Explore the Data

The dataset contains numerical input variables V1 to V28, which are the result of a PCA transformation of the original features (which could not be provided due to confidentiality issues).

The response variable `Class` tell us whether a transaction was fraudulent (value = 1) or not (value = 0).

```{r}
creditcard %>% ggplot(aes(x = Class)) + geom_bar(color = "grey", fill = "lightgrey") + theme_bw()
```

There are two additional features, `Time` (time in seconds between each transaction and the first transaction) and `Amount` (how much money was transferred in this transaction).

Because `Time` only tells us the order in which transactions have been done, it doesn’t actually tell us anything about the actual times (i.e. time of day) of the transaction. Therefore, I am normalizing them by day and bin them into four groups according to time of day.

```{r}
summary(creditcard$Time)
```

```{r}
# separate transactions by day
creditcard$day <- ifelse(creditcard$Time > 3600 * 24, "day2", "day1")#3600 = 1 hr = 60 mins = 60 x 60 s = 3600 s

# make transaction relative to day
creditcard$Time_day <- ifelse(creditcard$day == "day2", creditcard$Time - 86400, creditcard$Time)

summary(creditcard[creditcard$day == "day1", ]$Time_day)

summary(creditcard[creditcard$day == "day2", ]$Time_day)

# bin transactions according to time of day
creditcard$Time <- as.factor(ifelse(creditcard$Time_day <= 38138, "gr1", # mean 1st Qu.
                          ifelse(creditcard$Time_day <= 52327, "gr2", # mean mean
                                 ifelse(creditcard$Time_day <= 69580, "gr3", # mean 3rd Qu
                                        "gr4"))))

creditcard %>% ggplot(aes(x = day)) + geom_bar(color = "grey", fill = "lightgrey") + theme_bw()
```

We can see now that the transactions in this dataset have all been recorded on two consecutive days and there are roughly the same number of transactions on these two days.

Now, I remove the columns I used to create the Time bins.

```{r}
creditcard <- select(creditcard, -Time_day, -day)

# convert class variable to factor
creditcard$Class <- factor(creditcard$Class)

creditcard %>% ggplot(aes(x = Time)) + geom_bar(color = "grey", fill = "lightgrey") +
    theme_bw() + facet_wrap( ~ Class, scales = "free", ncol = 2)
```

The distribution of transactions over the four Time bins shows, that the majority of fraud cases have happened in group 1 (although, I can’t say when exactly because the original Time feature did not tell us when the first transaction occurred).

I also want to look at the distribution of the amounts of money that were transferred:

```{r}
summary(creditcard[creditcard$Class == "0", ]$Amount)

summary(creditcard[creditcard$Class == "1", ]$Amount)

creditcard %>% ggplot(aes(x = Amount)) + geom_histogram(color = "grey", fill = "lightgrey", bins = 50) +
    theme_bw() + facet_wrap( ~ Class, scales = "free", ncol = 2)
```

Interestingly, fraudulent credit card transactions had a higher mean amount of money that was transferred, but the maximum amount was much lower compared to regular transactions.

## H~2~O Modeling

```{r}
library(h2o)
h2o.init(nthreads = 2)#prefer -1 but need to process at work ;)

# convert data to H2OFrame
creditcard_hf <- as.h2o(creditcard)

splits <- h2o.splitFrame(creditcard_hf, ratios = c(0.4, 0.4), seed = 42)

train_unsupervised  <- splits[[1]]
train_supervised  <- splits[[2]]
test <- splits[[3]]

response <- "Class"
features <- setdiff(colnames(train_unsupervised), response)
```

## Auto-Encoder

Train the unsupervised neural network model using deep learning autoencoders. With h2o, simply set `autoencoder = TRUE`.

Applying a technique called “bottleneck” training, where the hidden layer in the middle is very small. This means the model will reduce the dimensionality of the input data (in this case, down to 2 nodes/dimensions).

The autoencoder model learns the patterns of the input data irrespective of given class labels. Here, it will learn, which credit card transactions are similar and which transactions are outliers or anomalies. We need to keep in mind the autoencoder models will be sensitive to outliers in our data, which might throw off otherwise typical patterns.

```{r eval=FALSE}
model_nn <- h2o.deeplearning(x = features, training_frame = train_unsupervised,
         model_id = "model_nn",
         autoencoder = TRUE,
         reproducible = TRUE, #slow - turn off for real problems
         ignore_const_cols = FALSE,
         seed = 42,
         hidden = c(10, 2, 10), 
         epochs = 100,
         activation = "Tanh")

#training can take a while -save the model:
h2o.saveModel(model_nn, path="C:/Users/czwea/Documents/GitHub/DeepLearning/Encoder/model_nn", force = TRUE)
```
```{r echo=FALSE}
model_nn <- h2o.loadModel("model_nn//model_nn")
model_nn
```

```{r}
#Convert to autoencoded representation
test_autoenc <- h2o.predict(model_nn, test)
```

### Dimensionality reduction with hidden layers

Because a bottleneck model with two nodes was used in the hidden layer in the middle, use this dimensionality reduction to explore the feature space (similar to what to we could do with a principal component analysis). We can extract this hidden feature with the `h2o.deepfeatures()` and plot it to show the reduced representation of the input data.

```{r}
train_features <- h2o.deepfeatures(model_nn, train_unsupervised, layer = 2) %>% as.data.frame() %>%
     mutate(Class = as.vector(train_unsupervised[, 31]))

ggplot(train_features, aes(x = DF.L2.C1, y = DF.L2.C2, color = Class)) + geom_point(alpha = 0.1)
```

We do not see a cluster of fraudulent transactions that is distinct from non-fraud instances, so dimensionality reduction with our autoencoder model alone is not sufficient to identify fraud in this dataset.

But we could use the reduced dimensionality representation of one of the hidden layers as features for model training. An example would be to use the 10 features from the first or third hidden layer:

```{r}
# let's take the third hidden layer
train_features <- h2o.deepfeatures(model_nn, train_unsupervised, layer = 3) %>% as.data.frame() %>%
     mutate(Class = as.factor(as.vector(train_unsupervised[, 31]))) %>% as.h2o()

features_dim <- setdiff(colnames(train_features), response)
```
```{r eval=FALSE}
model_nn_dim <- h2o.deeplearning(y = response, x = features_dim, training_frame = train_features,
                reproducible = TRUE, #slow - turn off for real problems
                balance_classes = TRUE,
                ignore_const_cols = FALSE,
                seed = 42,
                hidden = c(10, 2, 10), 
                epochs = 100,
                activation = "Tanh")

h2o.saveModel(model_nn_dim, path="C:/Users/czwea/Documents/GitHub/DeepLearning/Encoder/model_nn_dim", force = TRUE)
```
```{r echo=FALSE}
model_nn_dim <- h2o.loadModel("model_nn_dim/DeepLearning_model_R_1501157766634_1")#Check for correct model name
model_nn_dim
```

For measuring model performance on test data, we need to convert the test data to the same reduced dimensions as the trainings data:

```{r}
test_dim <- h2o.deepfeatures(model_nn, test, layer = 3)

h2o.predict(model_nn_dim, test_dim) %>% as.data.frame() %>% mutate(actual = as.vector(test[, 31])) %>%
     group_by(actual, predict) %>% summarise(n = n()) %>% mutate(freq = n / sum(n))
```

Now, this actually looks quite good in terms of identifying fraud cases: 92% (this time is was only 83%) of fraud cases were identified! However, many non-fraud cases were also classified as fraud. For real-life application, this wouldn’t be a good model. Let’s try some other techniques.

## Anomaly detection

We can also ask which instances were considered outliers or anomalies within our test data, using the `h2o.anomaly()` function. Based on the autoencoder model that was trained before, the input data will be reconstructed and for each instance, the mean squared error (MSE) between actual value and reconstruction is calculated.

Calculating the mean MSE for both class labels.

```{r}
anomaly <- h2o.anomaly(model_nn, test) %>% as.data.frame() %>% tibble::rownames_to_column() %>%
     mutate(Class = as.vector(test[, 31]))

mean_mse <- anomaly %>% group_by(Class) %>% summarise(mean = mean(Reconstruction.MSE))
```

Now plot.

```{r}
ggplot(anomaly, aes(x = as.numeric(rowname), y = Reconstruction.MSE, color = as.factor(Class))) +
  geom_point(alpha = 0.3) + geom_hline(data = mean_mse, aes(yintercept = mean, color = Class)) +
  scale_color_brewer(palette = "Set1") + labs(x = "instance number", color = "Class")
```

We can see in the plot, there is no perfect classification into fraud and non-fraud cases but the mean MSE is definitely higher for fraudulent transactions than for regular ones.

We can now identify outlier instances by applying an MSE threshold for what we consider outliers. We could e.g. say that we consider every instance with an MSE > 0.02 (chosen according to the plot above) to be an anomaly/outlier.

```{r}
anomaly <- anomaly %>% mutate(outlier = ifelse(Reconstruction.MSE > 0.02, "outlier", "no_outlier"))

anomaly %>% group_by(Class, outlier) %>% summarise(n = n()) %>% mutate(freq = n / sum(n)) 
```

Outlier detection is not sufficient to correctly classify fraudulent credit card transactions either (at least not with this dataset).

### Pre-trained supervised model

Try using the autoencoder model as a pre-training input for a supervised model. Here, I am again using a neural network. This model will now use the weights from the autoencoder for model fitting.

```{r eval=FALSE}
model_nn_2 <- h2o.deeplearning(y = response, x = features, training_frame = train_supervised,
                pretrained_autoencoder  = "model_nn",
                reproducible = TRUE, #slow - turn off for real problems
                balance_classes = TRUE,
                ignore_const_cols = FALSE,
                seed = 42,
                hidden = c(10, 2, 10), 
                epochs = 100,
                activation = "Tanh")

h2o.saveModel(model_nn_2, path="C:/Users/czwea/Documents/GitHub/DeepLearning/Encoder/model_nn_2", force = TRUE)
```
```{r echo=FALSE}
model_nn_2 <- h2o.loadModel("model_nn_2/DeepLearning_model_R_1501157766634_23")
model_nn_2
```

```{r}
pred <- as.data.frame(h2o.predict(object = model_nn_2, newdata = test)) %>% mutate(actual = as.vector(test[, 31]))

pred %>% group_by(actual, predict) %>% summarise(n = n()) %>% mutate(freq = n / sum(n)) 
```
```{r}
pred %>% ggplot(aes(x = actual, fill = predict)) + geom_bar() + theme_bw() +
     scale_fill_brewer(palette = "Set1") + facet_wrap( ~ actual, scales = "free", ncol = 2)
```

This looks much better! We did miss 17% of the fraud cases but we also did not mis-classify too many of the non-fraud cases.

In real-life, we would now spend some more time trying to improve the model by e.g. performing grid search for hyperparameter tuning, going back to the original features (which we did not have here) and trying different engineered features and/or trying different algorithms. But here, I will leave it at that.

## Measuring model performance on highly unbalanced data

Because of the severe bias towards non-fraud cases, we can not use performance measures like accuracy or area under the curve (AUC), as they would give overly optimistic results based on the high percentage of correct classifications of the majority class.

An alternative to AUC is to use the precision-recall curve or the sensitivity (recall)-specificity curve. To calculate and plot these metrics, we can use the `ROCR` package. There are different ways to calculate the area under a curve (see the `PRROC` package for details) but I am going to use a simple function that calculates the area between every consecutive points-pair of x (i.e. x1 - x0, x2 - x1, etc.) under the corresponding values of y.

```{r}
library(ROCR)

# http://stackoverflow.com/questions/24563061/computing-integral-of-a-line-plot-in-r
line_integral <- function(x, y) {
  dx <- diff(x)
  end <- length(y)
  my <- (y[1:(end - 1)] + y[2:end]) / 2
  sum(dx * my)
} 

prediction_obj <- prediction(pred$p1, pred$actual)

par(mfrow = c(1, 2))
par(mar = c(5.1,4.1,4.1,2.1))

# precision-recall curve
perf1 <- performance(prediction_obj, measure = "prec", x.measure = "rec") 

x <- perf1@x.values[[1]]
y <- perf1@y.values[[1]]
y[1] <- 0

plot(perf1, main = paste("Area Under the\nPrecision-Recall Curve:\n", round(abs(line_integral(x,y)), digits = 3)))

# sensitivity-specificity curve
perf2 <- performance(prediction_obj, measure = "sens", x.measure = "spec") 

x <- perf2@x.values[[1]]
y <- perf2@y.values[[1]]
y[1] <- 0

plot(perf2, main = paste("Area Under the\nSensitivity-Specificity Curve:\n", round(abs(line_integral(x,y)), digits = 3)))
```

Precision is the proportion of test cases predicted to be fraud that were indeed fraudulent (i.e. the true positive predictions), while recall or sensitivity is the proportion of fraud cases that were identified as fraud. And specificity is the proportion of non-fraud cases that are identified as non-fraud.

The precision-recall curve tells us the relationship between correct fraud predictions and the proportion of fraud cases that were detected (e.g. if all or most fraud cases were identified, we also have many non-fraud cases predicted as fraud and vice versa). The sensitivity-specificity curve thus tell us the relationship between correctly identified classes of both labels (e.g. if we have 100% correctly classified fraud cases, we will have no correctly classified non-fraud cases and vice versa).

We can also look at this a little bit differently, by manually going through different prediction thresholds and calculating how many cases were correctly classified in the two classes:

```{r}
thresholds <- seq(from = 0, to = 1, by = 0.1)
pred_thresholds <- data.frame(actual = pred$actual)

for (threshold in thresholds) {
  prediction <- ifelse(pred$p1 > threshold, 1, 0)
  prediction_true <- ifelse(pred_thresholds$actual == prediction, TRUE, FALSE)
  pred_thresholds <- cbind(pred_thresholds, prediction_true)
}

colnames(pred_thresholds)[-1] <- thresholds

pred_thresholds %>% gather(x, y, 2:ncol(pred_thresholds)) %>% group_by(actual, x, y) %>% summarise(n = n()) %>%
     ggplot(aes(x = as.numeric(x), y = n, color = actual)) + geom_vline(xintercept = 0.6, alpha = 0.5) +
     geom_line() + geom_point(alpha = 0.5) + theme_bw() + facet_wrap(actual ~ y, scales = "free", ncol = 2) +
     labs(x = "prediction threshold", y = "number of instances")
```

This plot tells us that we can increase the number of correctly classified non-fraud cases without loosing correctly classified fraud cases when we increase the prediction threshold from the default 0.5 to 0.6:

```{r}
pred %>% mutate(predict = ifelse(pred$p1 > 0.6, 1, 0)) %>% group_by(actual, predict) %>%
     summarise(n = n()) %>% mutate(freq = n / sum(n)) 
```

Our final model now correctly identified 83% of fraud cases and almost 100% of non-fraud cases.
```{r echo=FALSE}
h2o.shutdown(prompt = FALSE)
```







